---
title: "Algorithm benchmark in 10D"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(kldest)
library(ggplot2)
library(reshape2)
library(MASS)
set.seed(0)
```

In large dimensions, performance of 1-nearest-neighbour density estimation 
(one- or two-sample variants) decreases and the bias-reduced variant becomes 
important. Here, we evaluate a 10-D test case comparing correlated multivariate 
Gaussians.

## Specification of benchmark scenario

### Distributions and analytical KL divergence

Simulation study design

```{r}
D <- 10
Sigma <- constDiagMatrix(dim = D, diag = 1, offDiag = 0.999)
```


```{r}
p <- list(
    gaussian_10d = list(
        paramTrue   = list(mu = rep(0, D), sigma = Sigma),
        paramApprox = list(mu = rep(1, D), sigma = Sigma)
    )
)
distributions <- list(
    gaussian_10d = list(
        samples = function(n, m) {
            X <- mvrnorm(n = n, 
                         mu = p$gaussian_10d$paramTrue$mu,   
                         Sigma = p$gaussian_10d$paramTrue$sigma)
            Y <- mvrnorm(n = m, 
                         mu = p$gaussian_10d$paramApprox$mu, 
                         Sigma = p$gaussian_10d$paramApprox$sigma)
            list(X = X, Y = Y)
        },
        q = function(x) mvdnorm(x, mu = p$gaussian_10d$paramApprox$mu, 
                                Sigma = p$gaussian_10d$paramApprox$sigma),
        kld = kld_gaussian(
            mu1    = p$gaussian_10d$paramTrue$mu,
            sigma1 = p$gaussian_10d$paramTrue$sigma,
            mu2    = p$gaussian_10d$paramApprox$mu,
            sigma2 = p$gaussian_10d$paramApprox$sigma)
    )
)
```

Analytical values for Kullback-Leibler divergences in test cases:

```{r}
vapply(distributions, function(x) x$kld, 1)
```

### Simulation scenarios

For each of the distributions specified above, samples of different sizes are 
drawn, with several replicates per distribution and sample size.

```{r}
samplesize <- 10^(2:4)
#n <- c(20,50,100,200,500,1000,2000,5000,10000)
nRep       <- 25L

scenarios <- combinations(
    distribution = names(distributions),
    sample.size  = samplesize,
    replicate    = 1:nRep
)
```

### Algorithms

The following algorithms are considered:

```{r}
algorithms_XY <- list(
    nn1_XY   = function(X, Y) kld_est_nn(X, Y),
    nn_br_XY = function(X, Y) kld_est_brnn(X, Y, warn.max.k = FALSE)
)
algorithms_Xq <- list(
    nn1_Xq = function(X, q) kld_est_nn(X, q = q)
)
nAlgoXY  <- length(algorithms_XY)
nAlgoXq  <- length(algorithms_Xq)
nmAlgo   <- c(names(algorithms_XY),names(algorithms_Xq))
```

## Run the simulation study 

```{r}
# allocating results matrices
nscenario  <- nrow(scenarios)
runtime <- kld <- matrix(nrow = nscenario, 
                         ncol = nAlgoXY+nAlgoXq, 
                         dimnames = list(NULL, nmAlgo))

for (i in 1:nscenario) {

    dist <- scenarios$distribution[i]
    n    <- scenarios$sample.size[i]
    
    samples <- distributions[[dist]]$sample(n = n, m = n)
    X <- samples$X
    Y <- samples$Y
    q <- distributions[[dist]]$q

    # different algorithms are evaluated on the same samples
    for (j in 1:nAlgoXY) {
        algo         <- algorithms_XY[[j]]
        start_time   <- Sys.time()
        kld[i,j]     <- algo(X, Y)
        end_time     <- Sys.time()
        runtime[i,j] <- end_time - start_time
    }
    for (j in 1:nAlgoXq) {
        nj            <- nAlgoXY+j
        algo          <- algorithms_Xq[[j]]
        start_time    <- Sys.time()
        kld[i,nj]     <- algo(X, q)
        end_time      <- Sys.time()
        runtime[i,nj] <- end_time - start_time
    }
}
```


Post-processing: combine `scenarios`, `kldiv` and `runtime` into a single 
data frame

```{r}
tmp1 <- cbind(scenarios, kld) |> melt(measure.vars  = nmAlgo,
                                      value.name    = "kld",
                                      variable.name = "algorithm") 
tmp2 <- cbind(scenarios, runtime) |> melt(measure.vars  = nmAlgo,
                                          value.name    = "runtime",
                                          variable.name = "algorithm") 
results <- merge(tmp1,tmp2)
results$sample.size <- as.factor(results$sample.size)
rm(tmp1,tmp2)
```


## Graphical display of results

### Accuracy of KL divergence estimators

```{r, fig.width=7, fig.height=6}
ggplot(results, aes(x=sample.size, y=kld, color=algorithm)) + 
    geom_jitter(position=position_dodge(0.5)) + 
    facet_wrap("distribution", scales = "free_y") +
    geom_hline(data = data.frame(distribution = names(distributions), 
                                 kldtrue = vapply(distributions, function(x) x$kld,1)), 
               aes(yintercept = kldtrue)) +
    xlab("Sample sizes") + ylab("KL divergence estimate") + 
    ggtitle("Accuracy of different algorithms") +
    theme(plot.title = element_text(hjust = 0.5))
```

$\Rightarrow$ the bias-reduced nearest neighbour algorithm shows much better 
performance than either of the plain 1-nearest neighbour algorithms. In particular, 
even if the approximate density `q` is known, it may be preferable to simulate a 
(large) sample `Y` from $Q$ and use the bias-reduced algorithm on the two-sample
problem, rather than using `q`.


