---
title: "High-Dimensional Illustration of kldest"
author: "Niklas Hartung"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages

```{r setup}
library(kldest)
library(ggplot2)
library(MASS)
library(reshape2)
```

For reproducibility 

```{r seed}
set.seed(0)
```


## 2-D correlated example

```{r}
D <- 2
Sigma <- constDiagMatrix(dim = 10, diag = 1, offDiag = 0.99)

X <- mvrnorm(n = 100, mu = rep(0,D), Sigma = Sigma)
Y <- mvrnorm(n = 100, mu = rep(1,D), Sigma = Sigma)
```

Data in form of a matrix (i.i.d. samples across rows)

```{r}
head(X)
```

Plot the data

```{r}
plot(X[,1:2], xlim = c(-3,4),ylim = c(-3,4), 
     xlab = expression(x[1]), 
     ylab = expression(x[2]))
points(Y[,1:2], col = "red")
legend("topleft", 
       c("Samples from p","Samples from q"), 
       pch = 1, col = c("black","red"))
```

Analytical KL-D

```{r}
kld_gaussian(mu1 = rep(0,D), sigma1 = Sigma, mu2 = rep(1,D), sigma2 = Sigma)
```

Calculated KL-D

```{r}
kld_est(X, Y)
```

One-sample problem

```{r}
kld_est(X, q = function(x) mvdnorm(x, c(1,1), Sigma))
```

KL-D including a 95% confidence interval calculated by subsampling

```{r}
kld_ci_subsampling(X,Y)
```

One-sample problem

```{r}
kld_ci_subsampling(X, q = function(x) mvdnorm(x, c(1,1), Sigma))
```


## High-dimensional case

Simulation study design (high-dimensional correlated Gaussians)

```{r}
D <- 10
Sigma_10d <- constDiagMatrix(dim = D, diag = 1, offDiag = 0.999)
```


```{r}
paramTrue   <- list(mu = rep(0, D), sigma = Sigma)
paramApprox <- list(mu = rep(1, D), sigma = Sigma)
```


```{r}
sampler <- function(n, param) mvrnorm(n = n, 
                         mu = param$mu,   
                         Sigma = param$sigma)
```


```{r}
q <- function(x) mvdnorm(x, 
                         mu = paramApprox$mu, 
                         Sigma = paramApprox$sigma)
```

Analytical KL-D

```{r}
kld <- kld_gaussian(
    mu1    = paramTrue$mu,
    sigma1 = paramTrue$sigma,
    mu2    = paramApprox$mu,
    sigma2 = paramApprox$sigma
)
```

Samples of different sizes are drawn

```{r}
samplesize <- c(100,1000,10000)
```


Algorithms considered:


KL divergence estimation

```{r}

data.frame()
kld_ <- matrix(NA, nrow = 3, ncol = 3)

```



Plotting

```{r}

```















